\documentclass[11pt]{article}

% ---------------------------------------------------------
% Packages
% ---------------------------------------------------------
\usepackage[a4paper, margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{physics}
\usepackage{bm}
\usepackage{listings}
\usepackage{xcolor}


% ---------------------------------------------------------
% Code block style
% ---------------------------------------------------------
\lstset{
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue},
    commentstyle=\color{gray},
    stringstyle=\color{orange},
    breaklines=true,
    columns=fullflexible
}

% ---------------------------------------------------------
% indentation
% ---------------------------------------------------------
\setlength{\parindent}{0pt}


% ---------------------------------------------------------
% Hyperlink setup
% ---------------------------------------------------------
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    urlcolor=blue,
    citecolor=blue
}

% ---------------------------------------------------------
% Title
% ---------------------------------------------------------
\title{
    \vspace{-1cm}
    \textbf{Lab 1 Report}\\
    \vspace{0.2cm}
    Summer 2026
}
\author{
    Soeun Jang \\
    Virginia Tech \\
    \texttt{sophiaj04@vt.edu}
}
\date{\today}

% ---------------------------------------------------------
% Document
% ---------------------------------------------------------
\begin{document}

\maketitle
\newpage   

\tableofcontents
\newpage   

% ---------------------------------------------------------
\section*{Introduction}
\addcontentsline{toc}{section}{Introduction}
% ---------------------------------------------------------

In this lab, we focus on representing matrices in low-rank form when solving PDEs. 
Using truncated SVD significantly reduces memory usage and computational cost. The goal of this lab is to study how the rank evolves over time and how different truncation strategies, together with Taylor-series time stepping and the LLRSVD structure, affect the accuracy and efficiency of the solution.


% ---------------------------------------------------------
\section*{Part 1: The LLRSVD and Truncation}
\addcontentsline{toc}{section}{Part 1: The LLRSVD and Truncation}
% ---------------------------------------------------------

\subsection*{Exercise 1}
\textbf{1}.Defining the following mutable structure in Julia:
\begin{lstlisting}
mutable struct LLRSVD
    U::Matrix{Float64}
    S::Vector{Float64}
    V::Matrix{Float64}
    m::Int
    n::Int
    r::Int
end
\end{lstlisting}

In this exercise, we define a mutable structure called \textbf{LLRSVD} that stores a matrix in low-rank SVD form. 
The structure contains the left singular vectors \textbf{U}, the singular values \textbf{S}, and the right singular vectors \textbf{V}. 
The fields \textbf{m},\textbf{n} and \textbf{r} keep track of the original matrix dimensions and the current rank of the low-rank representation.

\vspace{1.5em}

\textbf{2}. Convenience Constructor

\begin{lstlisting}
function LLRSVD(U::Matrix{Float64}, S::Vector{Float64}, V::Matrix{Float64})
    m, r1 = size(U)
    r2 = length(S)
    n, r3 = size(V)
    @assert r1 == r2 == r3
    return LLRSVD(U, S, V, m, n, r1)
end
\end{lstlisting}

This constructor builds an LLRSVD object directly from pre-computed SVD factors. For the SVD to be valid, the ranks inferred from U, S, and V (r1, r2, r3) must all be equal. After asserting this consistency, the constructor automatically infers the dimensions m, n, and the rank r, and then returns a properly structured LLRSVD object.

\vspace{1.5em}

\textbf{3}. Constructor W

\begin{lstlisting}
function LLRSVD(W::Matrix{Float64}, TOL::Float64)
    F = svd(W)
    U, S, V = F.U, F.S, F.V
    s1 = S[1]
    keep = findall(s -> s >= TOL * s1, S)

    if isempty(keep)
        keep = [1]
    end

    return LLRSVD(U[:, keep], S[keep], V[:, keep],
                  size(W,1), size(W,2), length(keep))
end
\end{lstlisting}

In Julia, the SVD of a matrix W is returned in the form 
$W = U \cdot \mathrm{diag}(S) \cdot V^T$.  Since the LLRSVD structure stores the factors as U, S, and V, we would need to convert the transpose of V back to V by taking another transpose. Using the structured form F = svd(W) avoids this problem of the code because F.V already gives the correct V without requiring any additional transposes. This helps the constructor to be more clear and reduce confusion.

\vspace{1.5em}

\textbf{4}. todense 

\begin{lstlisting}
module ToDenseModule

using LinearAlgebra
import ..LLRSVDModule: LLRSVD

export todense

function todense(A::LLRSVD)::Matrix{Float64}
    return A.U * Diagonal(A.S) * A.V'
end

end 

\end{lstlisting}

The todense function reconstructs the full matrix from its low-rank SVD stored in LLRSVD. Since the structure already contains U,S, and V, the dense matrix can be computed simply by multiplying those components.

\vspace{1.5em}

\textbf{5}. toelem

\begin{lstlisting}
module ToElemModule

import ..LLRSVDModule: LLRSVD

export toelem

function toelem(A::LLRSVD, i::Int, j::Int)::Float64
    return dot(A.U[i, :], A.S .* A.V[j, :])
end

end
\end{lstlisting}

The toelem function reconstructs $A(i,j)$ from the stored low-rank SVD. Since $A \approx USV^T$, the entry of i and j can be written as a dot product between the $i$th row of U and the vector with $k$th entry for S and V. This leads to the implementation of \textbf{dot(A.U[i, :], A.S .* A.V[j, :])}.
\vspace{1.5em}

\textbf{6}.Verification

What we need to verify is for a random 30 $\times$ 20 matrix 

\[
\| W - \text{todense}(\text{LLRSVD}(W, 0.0)) \| < 10^{-12}.
\]



\begin{lstlisting}
using Test
using LinearAlgebra
include("../src/LLRSVD.jl")
include("../src/todense.jl")

using .LLRSVDModule: LLRSVD
using .ToDenseModule: todense

@testset "LLRSVD Exercise 1" begin
    W = randn(30, 20)
    A = LLRSVD(W, 0.0)
    err = norm(W - todense(A))
    @test err < 1e-12
end
\end{lstlisting}

The implementation of the \texttt{LLRSVD} data structure and its associated
functions was validated through a reconstruction test on a random matrix.
Using the \texttt{todense} function, we obtained a reconstruction error of
approximately $4.145 \times 10^{-14}$, which indicates that the low-rank SVD
representation was implemented correctly and reproduces the original matrix to
machine precision. The test returned \texttt{true}, confirming that our
constructor behaves as expected and that the LLRSVD framework is functioning
properly.

\subsection*{Exercise 2}

\textbf{1}. Implement truncation

\begin{lstlisting}
function truncate(A::LLRSVD, TOL::Float64)::LLRSVD
    S = A.S
    t = sum(S.^2)
    k = 0.0
    r = 0

    for i in 1:length(S)
        k += S[i]^2
        tail = t - k
        if tail <= TOL
            r = i
            break
        end
    end

    if r == 0
        r = length(S)
    end

    return LLRSVD(A.U[:,1:r], A.S[1:r], A.V[:,1:r], A.m, A.n, r)
end
\end{lstlisting}

By implementing Frobenius-norm truncation, we use that the squared norm of the error is 
\[
\|A - A_r\|_F^2 = \sum_{i=r+1}^n \sigma_i^2.
\]
The smallest rank $r$ is chosen so that the tail energy 
$\sum_{i=r+1}^n \sigma_i^2$ is less than or equal to the tolerance TOL. 
The algorithm accumulates the squared singular values until this condition 
is met and returns the corresponding low-rank LLRSVD approximation.

\vspace{1.5em}

\textbf{2}. Verification
\begin{lstlisting}
using Test
using LinearAlgebra
include("../src/LLRSVD.jl")
include("../src/todense.jl")
include("../src/toelem.jl")

using .LLRSVDModule: LLRSVD, truncate
using .ToDenseModule: todense

@testset "Exercise 2: Truncation Verification" begin
    W = randn(50, 40)
    A = LLRSVD(W, 0.0)

    tols = [1e-1, 1e-3, 1e-6, 1e-10]

    for TOL in tols
        A2 = LLRSVDModule.truncate(A, TOL)
        err = norm(todense(A) - todense(A2))
        @test err <= TOL
    end
end
\end{lstlisting}

For this verification, a random matrix was generated using
\texttt{randn}. The singular values of this matrix remained relatively
large, so the truncation procedure retained essentially all singular
values. As a result, the truncated matrix was identical to the original
LLRSVD representation, producing zero reconstruction error for all
tested tolerances. This outcome confirms that the truncation function correctly preserves the
original matrix when the tolerance is sufficiently small, validating the implementation of the truncation algorithm.


\vspace{1.5em}
\textbf{3}. The reason why the loop must terminate with r is equal or greater than 1 for any TOL that is greater than 0.
\vspace{1.5em}
Since the singular values are greater than 0 and ordered decreasingly, the truncation algorithm always keeps at least the largest singular value. If all singular values are removed, the truncation error would be equal to the Frobenius norm of the original matrix which is not satisfying the small positive tolerance. Therefore, for any $TOL > 0$, the algorithm must terminate
with $r \geq 1$.

% ---------------------------------------------------------
\section*{Part 2: Taylor Series and Truncation Tolerances}
\addcontentsline{toc}{section}{Part 2: Taylor Series and Truncation Tolerances}
% ---------------------------------------------------------
\subsection*{Exercise 3}
\textbf{1}. Construct grid vectors

\begin{lstlisting}
function grid()
    m = 64
    n = 64

    x = range(0, 2*pi, length=m+1)[1:end-1]
    y = range(0, 2*pi, length=n+1)[1:end-1]

    u_x = sin.(x)
    u_y = sin.(y)

    W0 = u_x * u_y'

    return x, y, W0
end

\end{lstlisting}
The initial condition is defined as
\[
u_0(x,y) = \sin(x)\sin(y),
\]
that leads to a separable representaion, which means that the initial matrix can be written as an outer product.
\[
W^{(0)} = \sin(x)\sin(y)^T,
\]
which indicates that W contains the rank 1.

\vspace{1.5em}
\textbf{2}. Second-order centered finite difference Matrices

\begin{lstlisting}
function D(N,h)
    L = N * h
    scale = 2*pi / L
    p = zeros(Float64, N, N)

    for i in 1:N
        for j in 1:N
            if i != j
                p[i, j] = 0.5 * (-1)^(i - j) * cot(pi * (i - j) / N) * scale
            end
        end
    end

    return p
end

function diff(x, y)
    Dx = D(length(x), step(x))
    Dy = D(length(y), step(y))
    return Dx, Dy
end

\end{lstlisting}

We construct the discrete first derivative operators $D_x$ and $D_y$ using second order centered finite difference matrices on a periodic grid. 
The first derivative is
\[
(Du)_i = \frac{u_{i+1} - u_{i-1}}{2h},
\]
which is second order. These matrices provide second order approximations or x and y for periodic functions.

\textbf{3}. Verification

\begin{lstlisting}
using Test
using LinearAlgebra

include("../src/grid.jl")
include("../src/secondOrder.jl")

@testset "Exercise 3" begin
  x, y, W0 = grid()
  Dx, Dy = diff(x, y)

  Wx_num = Dx * W0
  Wy_num = W0 * Dy'

  Wx_exact = cos.(x) * sin.(y)'
  Wy_exact = sin.(x) * cos.(y)'

  R = Wx_num + Wy_num - Wx_exact - Wy_exact

  err_F = norm(R)

  @test err_F < 1e-10
end
\end{lstlisting}

To confirm that the $D_x$ and $D_y$ are implemented well, we are using numerical derivatives of W from its initial condition. We brought out its exact derivatives for each x and y and computed the Frobenius norm of the difference between numerical and exact derivation. It passed the test, which means its norm is extremely small. 

\subsection*{Exercise 4}
\textbf{1}. Compute Tk and Plot singular values

\begin{lstlisting}
x, y, W0 = grid()
Dx, Dy = diff(x, y)
\end{lstlisting}
This is to set up the initial matrix. 
\begin{lstlisting}
function dense(W, Dx, Dy)
    Dx*W + W*Dy'
end

Wk = [W0]   # W0 is W^(0)

for k in 1:7
    push!(Wk, dense(Wk[end], Dx, Dy))
end
\end{lstlisting}

To apply the derivative operator, we need this step to get values:
Wk[1] = W$^{0}$, Wk[2] = W$^{1}$ ... and so on. 

We have to convert it to taylor terms.

\begin{lstlisting}
dt = 0.1
Tk = [ (dt^k / factorial(k)) * Wk[k+1] for k in 0:7 ]
\end{lstlisting}

Now it became Tk[1] = T$^{0}$, Tk[2] = T$^{1}$ ... and so on.

Now we are computing singular values and plot them.

\begin{lstlisting}
#plot the singular values
    p = plot(yscale=:log10, legend=:outertopright)

    for k in 1:8
        plot!(p, sv[k], label="T$(k-1)")
    end

    display(p)
\end{lstlisting}
\vspace{1.5em}
\textbf{2}. Tk recording

\begin{lstlisting}
sv = [ svd(Tk[k+1]).S for k in 0:7 ]
norms = [ sqrt(sum(sv[k+1].^2)) for k in 0:7 ]
println(norms)
\end{lstlisting}

To record this, let's do test.
\begin{lstlisting}
@testset "Exercise 4-2" begin
    x, y, W0 = grid()
    Dx, Dy = diff(x, y)

    dense(W, Dx, Dy) = Dx * W + W * Dy'
    Wk = [W0]
    for k in 1:7
        push!(Wk, dense(Wk[end], Dx, Dy))
    end

    dt = 0.1
    Tk = [ (dt^k / factorial(k)) * Wk[k+1] for k in 0:7 ]

    sv = [ svd(Tk[k+1]).S for k in 0:7 ]

    norms = [ sqrt(sum(sv[k+1].^2)) for k in 0:7 ]

    @test length(norms) == 8

    global TK_NORMS = norms
end
\end{lstlisting}

Based on the taylor terms, the result norms \textbf{decrease rapidly} as k increases. Since the factorial in the denominator grows much faster than numerator, high-order taylor terms become very little to the final sum.
\vspace{1.5em}

\textbf{3}. how to choose TOL

A taylor method of order p has local truncation error $O((\Delta t)^{p+1})$. This means the error should not be greater than this when we divide each $T_k$ into SVD. $T_k$'s each size is decreasing rapidly as k increases, which means we can allow to use larger TOL when k is large. Therefore, choosing TOL as 
\[
TOL(k) \approx (\Delta t)^{p+1-k}
\]
so the sum of truncation error is always smaller than taylor truncation error. 
To verify this, we need a code.

\begin{lstlisting}
@testset "Exercise 4-3" begin
    x, y, W0 = grid()
    Dx, Dy = diff(x, y)
    dense(W, Dx, Dy) = Dx * W + W * Dy'

    Wk = [W0]
    for k in 1:7
        push!(Wk, dense(Wk[end], Dx, Dy))
    end

    dt = 0.1
    p = 7
    Wref = [ sin(x[i] - dt) * sin(y[j] - dt) for i in 1:length(x), j in 1:length(y) ]


    alphas = [0.0, 1e-3, 1e-2, 1e-1, 1.0, 10.0]
    results = Dict{Float64,Float64}()

    for alpha in alphas
        Tk_trunc = []
        for k in 0:p
            Tk = (dt^k / factorial(k)) * Wk[k+1]
            U, S, V = svd(Tk)
            if alpha == 0.0
                r = length(S)
            else
                tol = alpha * dt^(p+1-k)
                r = sum(S .> tol)
                if r == 0
                    r = 1
                end
            end
            push!(Tk_trunc, U[:,1:r] * Diagonal(S[1:r]) * V[:,1:r]')
        end

        W_TS = sum(Tk_trunc)
        err = norm(vec(W_TS - Wref))
        println("alpha=", alpha, "  err=", err)
        results[alpha] = err
    end

    global ERR_43 = results
    
    minerr = minimum(values(results))
    @show minerr
    @test minerr < 10.0
end
  
\end{lstlisting}
This tolerance rule successfully keeps the truncation erro below the taylor truncation error, test has been passed. Therefore, it confirms that the SVD truncation does not significantly affect the method.

% ---------------------------------------------------------
\section*{Conclusion}
\addcontentsline{toc}{section}{Conclusion}
% ---------------------------------------------------------


\end{document}
