\documentclass[11pt]{report}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{float}
\usepackage{hyperref}
\usepackage{bm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{tocloft}
\usepackage{enumitem}

\title{%
    \textbf{Low-Rank and Tensor Network Methods for Differential Equations}\\[0.5em]
    final report
}
\author{Soeun Jang \\ Virginia Tech \\ \texttt{sophiaj04@vt.edu}}
\date{June 2026}

\begin{document}

\maketitle
\tableofcontents

\newpage


% =========================================================
\chapter{Lab 1: SVD, Taylor Series, and Low-Rank Truncation}

\section{Objectives}

This chapter helps to understand:
\begin{itemize}[leftmargin=*]
    \item Implement the LLRSVD data structure and use it to represent matrices in a compressed low-rank format.
    \item Truncate a low-rank representation to a prescribed Frobenius-norm tolerance without recomputing the SVD.
    \item Add a list of low-rank matrices and recompress the result in a single pass.
    \item Explain how the Taylor series order \(p\) determines the per-term SVD truncation tolerances, and verify this experimentally.
    \item Build a Step-and-Truncate time-stepper for a 2D PDE that keeps the solution in low-rank form throughout.
    \item Apply the solver to solid-body rotation and study how the effective rank evolves in time.
\end{itemize}

\section{Background}

Consider the two-dimensional constant-coefficient advection equation
\begin{equation}
    u_t = u_x + u_y, \qquad (x,y) \in [0,2\pi]^2,\ t>0,
\end{equation}
with periodic boundary conditions and initial condition \(u_0(x,y) = u(x,y,0)\).
The exact solution is the rigid translation
\begin{equation}
    u(x,y,t) = u_0(x+t, y+t).
\end{equation}

\subsection{Spatial Discretization}

The domain is discretized on an \(m \times n\) uniform mesh with grid points


\[
x_i = (i-1)h_x,\quad y_j = (j-1)h_y,\quad h_x = \frac{2\pi}{m},\quad h_y = \frac{2\pi}{n}.
\]


Grid values \(W_{ij} \approx u(x_i,y_j,t)\) are stored as an \(m \times n\) matrix \(W\).
Differentiation matrices \(D_x \in \mathbb{R}^{m \times m}\) and \(D_y \in \mathbb{R}^{n \times n}\) satisfy


\[
\partial_x W \approx D_x W,\qquad \partial_y W \approx W D_y^T.
\]


Define the semidiscrete operator


\[
\mathcal{L}(W) = D_x W + W D_y^T.
\]



\subsection{Taylor Series Time-Stepping}

Advancing from \(t=0\) by one step \(\Delta t\) via a Taylor series of order \(p\):
\begin{align}
    W^{(0)} &= W(0),\\
    W^{(k)} &= \mathcal{L}\big(W^{(k-1)}\big),\qquad k=1,\dots,p,\\
    W(\Delta t) &\approx \sum_{k=0}^{p} \frac{(\Delta t)^k}{k!} W^{(k)}.
\end{align}

\section{Part 1: LLRSVD Type and Truncation}

\subsection{Low-Rank SVD Representation}

A low-rank SVD representation stores a matrix \(A \approx U \Sigma V^T\) through three factors:
\begin{itemize}[leftmargin=*]
    \item \(U \in \mathbb{R}^{m \times r}\) with orthonormal columns,
    \item \(\Sigma = \mathrm{diag}(\sigma_1,\dots,\sigma_r)\) with \(\sigma_1 \ge \dots \ge \sigma_r > 0\),
    \item \(V \in \mathbb{R}^{n \times r}\) with orthonormal columns.
\end{itemize}
The dense matrix is recovered as \(U \mathrm{diag}(S) V^T\), where \(S\) is the vector of singular values.

\subsection{LLRSVD Struct (Conceptual)}

In Julia, one defines:
\begin{verbatim}
mutable struct LLRSVD
    U :: Matrix{Float64}  # m x r, orthonormal columns
    S :: Vector{Float64}  # length r, singular values (descending)
    V :: Matrix{Float64}  # n x r, orthonormal columns
    m :: Int
    n :: Int
    r :: Int
end
\end{verbatim}

\subsection{Frobenius-Norm Truncation}

Given singular values \(\sigma_1 \ge \dots \ge \sigma_R\), seek the smallest rank \(r\) such that
\begin{equation}
    \left\| A - A_r \right\|_F^2
    = \sum_{i=r+1}^{R} \sigma_i^2 \le \mathrm{TOL},
\end{equation}
where \(A_r\) is the truncated rank-\(r\) approximation.

\section{Part 2: Taylor Series and Truncation Tolerance}

\subsection{Full-Rank Example}

For \(m=n=64\), consider the initial condition


\[
u_0(x,y) = \sin(x)\sin(y),
\]


which has rank 1 as a matrix: \(W(0) = \sin(x)\sin(y)^T\).

Constructing \(D_x, D_y\) and verify that


\[
\left\| D_x W(0) + W(0) D_y^T - \cos(x)\sin(y)^T - \sin(x)\cos(y)^T \right\|_F
\]


is small.

\subsection{Singular Values of Taylor Terms}

For a fixed \(\Delta t\), compute the Taylor terms


\[
T_k = \frac{(\Delta t)^k}{k!} W^{(k)},\qquad k=0,\dots,p.
\]


For each \(k\), compute the SVD of \(T_k\) and study:
\begin{itemize}[leftmargin=*]
    \item decay of singular values,
    \item decay of \(\|T_k\|_F\) with \(k\),
    \item choice of per-term truncation tolerance so that truncation error does not dominate the Taylor truncation error \(O((\Delta t)^{p+1})\).
\end{itemize}

\section{Part 3: Adding Low-Rank Matrices (trunc\_sum)}

Implementing a function \texttt{trunc\_sum(terms, TOL)} that:
\begin{itemize}[leftmargin=*]
    \item sums a list of LLRSVD matrices,
    \item recompresses the result to tolerance \(\mathrm{TOL}\),
    \item avoids forming the full \(m \times n\) matrix.
\end{itemize}

\section{Part 4: Step-and-Truncate Simulation}

Building a full Step-and-Truncate solver for the 2D transport equation, comparing:
\begin{itemize}[leftmargin=*]
    \item Strategy A: truncate each Taylor term first, then sum;
    \item Strategy B: sum all terms, then truncate once.
\end{itemize}

\section{Part 5: Solid-Body Rotation}

For the PDE


\[
u_t = y u_x - x u_y,\qquad (x,y)\in[-6,6]^2,
\]


with velocity field \(v=(y,-x)\), the exact solution is


\[
u(x,y,t) = u_0(x\cos t + y\sin t,\ -x\sin t + y\cos t).
\]


Extending the low-rank machinery to this variable-coefficient case and studying rank evolution and error over one full revolution.

% =========================================================
\chapter{Lab 2: Low-Rank Sensitivities and Gradient Computation}

\section{Objectives}

This chapter helps to understand:
\begin{itemize}[leftmargin=*]
    \item Derive the matrix sensitivity equation for a parameter-dependent hyperbolic PDE.
    \item Implement a dense reference solver for the joint \((u,X)\) system.
    \item Diagnose compressibility of \(X(t)\) via its singular spectrum and numerical rank.
    \item Build a fixed-rank step-and-truncate RK4 integrator for \(u\) and \(X \approx U S V^T\).
    \item Evaluate the gradient \(\nabla_\alpha J\) directly from low-rank factors at \(O((n+p)r)\) cost.
    \item Use the cheap gradient in a steepest-descent loop with Armijo line search.
\end{itemize}

\section{Background: Parameter-Dependent PDE}

Consider
\begin{equation}
    u_t + A(x) u_x = 0,\qquad x\in[-5,5)\ \text{(periodic)},\quad 0<t<T,\quad u(x,0)=u_0(x),
\end{equation}
where \(A(x)\) is the wave speed. After discretization on \(n\) points, the state \(u(t)\in\mathbb{R}^n\) and parameters


\[
\alpha = [A(x_1),\dots,A(x_p)] \in \mathbb{R}^{1\times p},\quad p=n.
\]



\subsection{Cost Functional and Sensitivity Matrix}

Define


\[
J(u(T);u_0) = \frac{1}{2}\|u(T)-u_0\|_2^2.
\]


By the chain rule,


\[
\frac{\partial J}{\partial \alpha}
= (u(T)-u_0)^T \frac{\partial u(T)}{\partial \alpha}
= (u(T)-u_0)^T X(T),
\]


where \(X(t) := \partial u(t)/\partial \alpha \in \mathbb{R}^{n\times p}\).

\section{Semi-Discrete System and Sensitivity Equation}

For the variable-coefficient advection example,


\[
\frac{du}{dt} = -\mathrm{diag}(a)\, (D u),\qquad u(0)=u_0,
\]


with \(a_j = a(x_j)\). The sensitivity matrix \(X\) satisfies


\[
\frac{dX}{dt} = -\mathrm{diag}(D u) - \mathrm{diag}(a)\, D X,\qquad X(0)=0.
\]



\section{Dense Reference Solver}

Implementing a dense RK4 solver for the coupled \((u,X)\) system:
\begin{itemize}[leftmargin=*]
    \item CFL-based time step \(\Delta t = h / \max_j |a_j|\),
    \item total time \(T=10\),
    \item track singular values \(\sigma_k(X(t))\) and numerical rank
    

\[
    r_\varepsilon(t) = \min\{k : \sigma_{k+1}(X(t)) \le \varepsilon\, \sigma_1(X(t))\}
    \]


    for various \(\varepsilon\).
\end{itemize}

\section{Step-and-Truncate RK4 for Low-Rank \(X\)}

Represent \(X \approx U S V^T\) with \(U,V\in\mathbb{R}^{N\times r}\), \(S\in\mathbb{R}^{r\times r}\), and propagate:
\begin{itemize}[leftmargin=*]
    \item advance \(u\) with RK4,
    \item at each stage, update low-rank factors for \(X\) using:
    

\[
    \frac{dX}{dt} = -\mathrm{diag}(D u) - \mathrm{diag}(a)\, D X,
    \]


    \item concatenate factors, perform thin QR and small SVD,
    \item truncate back to fixed rank \(r\).
\end{itemize}

Compare:
\begin{itemize}[leftmargin=*]
    \item wall-clock time per step vs. dense solver,
    \item relative error \(\|X_r(T)-X_{\text{dense}}(T)\|_F / \|X_{\text{dense}}(T)\|_F\),
    \item error growth vs. time for different \(r\).
\end{itemize}

\section{Gradient from Low-Rank Factors}

If \(X(T) \approx U S V^T\), then


\[
\nabla_\alpha J(T)
= X(T)^T (u(T)-u_0)
= V \left( S^T \left( U^T (u(T)-u_0) \right) \right),
\]


which costs \(O((N+p)r)\) flops and never forms \(X(T)\) explicitly.

\section{Steepest Descent with Armijo Line Search}

At iteration \(k\), with parameters \(\alpha_k\):
\begin{itemize}[leftmargin=*]
    \item search direction: \(d_k = -\nabla_\alpha J(\alpha_k)\),
    \item step length \(\eta_k\) chosen by backtracking:
    

\[
    J(\alpha_k + \eta_k d_k)
    \le J(\alpha_k) + c_1 \eta_k \nabla_\alpha J(\alpha_k)^T d_k,\quad c_1\in(0,1),
    \]


    \item start from \(\eta_0\) (e.g. 1) and reduce by factor \(p\in(0,1)\) until Armijo condition holds.
\end{itemize}

% =========================================================
\chapter{Lab 3: Tucker Tensors}

\section{Objectives}

This chapter helps to understand:
\begin{itemize}[leftmargin=*]
    \item Compute the mode-\(n\) unfolding (matricization) of a 3-way tensor.
    \item Apply the mode-\(n\) product to contract a tensor with a matrix along a given mode.
    \item Implement the Higher-Order SVD (HOSVD) to obtain a Tucker decomposition.
    \item Truncate a Tucker decomposition to a prescribed multilinear rank and quantify the approximation error.
    \item Represent a 3D function on a fine grid in compressed Tucker format.
    \item Perform addition, inner product, and reconstruction directly in Tucker format.
\end{itemize}

\section{Tensors and Mode-\(n\) Unfolding}

A 3-way tensor \(A \in \mathbb{R}^{n_1 \times n_2 \times n_3}\) has entries \(a_{i_1 i_2 i_3}\).
The mode-\(n\) unfolding \(A^{(n)}\) arranges mode-\(n\) fibers as columns:


\[
A^{(1)} \in \mathbb{R}^{n_1 \times (n_2 n_3)},\quad
A^{(2)} \in \mathbb{R}^{n_2 \times (n_1 n_3)},\quad
A^{(3)} \in \mathbb{R}^{n_3 \times (n_1 n_2)}.
\]



\section{Mode-\(n\) Product}

Given \(A \in \mathbb{R}^{n_1 \times n_2 \times n_3}\) and \(M \in \mathbb{R}^{s \times n_n}\), the mode-\(n\) product
\(B = A \times_n M\) has entries


\[
b_{i_1 \dots i_{n-1} j i_{n+1} \dots i_3}
= \sum_{i_n=1}^{n_n} a_{i_1 \dots i_n \dots i_3} M_{j,i_n},
\]


and satisfies


\[
B^{(n)} = M A^{(n)}.
\]


Mode products along different modes commute:


\[
(A \times_m M) \times_n N = (A \times_n N) \times_m M,\quad m\neq n.
\]



\section{Tucker Decomposition and HOSVD}

A Tucker decomposition of \(A\) with multilinear rank \((r_1,r_2,r_3)\) is


\[
A \approx G \times_1 U^{(1)} \times_2 U^{(2)} \times_3 U^{(3)},
\]


where:
\begin{itemize}[leftmargin=*]
    \item \(G \in \mathbb{R}^{r_1 \times r_2 \times r_3}\) is the core tensor,
    \item \(U^{(n)} \in \mathbb{R}^{n_n \times r_n}\) have orthonormal columns.
\end{itemize}

\subsection{HOSVD Algorithm}

\begin{enumerate}[leftmargin=*]
    \item For \(n=1,2,3\): compute thin SVD of \(A^{(n)} = \widehat{U}^{(n)} \widehat{\Sigma}^{(n)} \widehat{V}^{(n)T}\).
          Retain leading \(r_n\) left singular vectors as \(U^{(n)}\).
    \item Compute core:
    

\[
    G = A \times_1 U^{(1)T} \times_2 U^{(2)T} \times_3 U^{(3)T}.
    \]


\end{enumerate}

The HOSVD error satisfies


\[
\left\| A - G \times_1 U^{(1)} \times_2 U^{(2)} \times_3 U^{(3)} \right\|_F^2
\le \sum_{n=1}^{3} \sum_{j>r_n} \sigma_j\big(A^{(n)}\big)^2,
\]


which is at most \(3\) times the best rank-\((r_1,r_2,r_3)\) approximation error.

\section{Tucker3 Struct and Reconstruction}

Conceptually, in Julia:
\begin{verbatim}
mutable struct Tucker3
    G  :: Array{Float64,3}  # r1 x r2 x r3 core
    U1 :: Matrix{Float64}   # n1 x r1
    U2 :: Matrix{Float64}   # n2 x r2
    U3 :: Matrix{Float64}   # n3 x r3
end
\end{verbatim}

Reconstruction:


\[
A \approx G \times_1 U_1 \times_2 U_2 \times_3 U_3.
\]



\section{Tucker Rounding}

Given an existing Tucker decomposition \((G,U_1,U_2,U_3)\), Tucker rounding recompresses the core:
\begin{itemize}[leftmargin=*]
    \item apply HOSVD to \(G\),
    \item obtain new core \(\widetilde{G}\) and mode matrices \(\widehat{U}^{(n)}\),
    \item set \(U_n^{\text{new}} = U_n \widehat{U}^{(n)}\).
\end{itemize}
This avoids expanding to the full \(n_1 \times n_2 \times n_3\) tensor.

\section{Inner Product and Norm in Tucker Format}

For two Tucker tensors


\[
A = G \times_1 U^{(1)} \times_2 U^{(2)} \times_3 U^{(3)},\quad
B = H \times_1 W^{(1)} \times_2 W^{(2)} \times_3 W^{(3)},
\]


the Frobenius inner product can be written as


\[
\langle A,B\rangle_F
= \sum_{j_1,j_2,j_3} \sum_{k_1,k_2,k_3}
    g_{j_1 j_2 j_3} h_{k_1 k_2 k_3}
    (U^{(1)T} W^{(1)})_{j_1 k_1}
    (U^{(2)T} W^{(2)})_{j_2 k_2}
    (U^{(3)T} W^{(3)})_{j_3 k_3},
\]


which costs \(O(r^6 + n r^2)\) instead of \(O(n^3)\).
The norm is \(\|A\|_F = \sqrt{\langle A,A\rangle_F}\).

\section{Sum and Recompression}

The sum \(A+B\) of two Tucker tensors with ranks \((r_1^A,r_2^A,r_3^A)\) and \((r_1^B,r_2^B,r_3^B)\) has a Tucker representation with ranks


\[
(r_1^A + r_1^B,\ r_2^A + r_2^B,\ r_3^A + r_3^B),
\]


and factor matrices formed by column concatenation \([U^{(n)}_A\ | U^{(n)}_B]\).
The combined core is block-diagonal in a suitable sense and can be recompressed via Tucker rounding.

\section{3D Advection in Tucker Format}

For


\[
u_t = u_x + u_y + u_z,\qquad (x,y,z)\in[0,2\pi]^3,\quad u_0(x,y,z)=\sin x\,\sin y\,\sin z,
\]


the exact solution is


\[
u(x,y,z,t) = u_0(x-t,y-t,z-t).
\]


The semidiscrete operator is


\[
\mathcal{L}(W) = D_x \times_1 W + D_y \times_2 W + D_z \times_3 W.
\]



If \(W\) is Tucker with rank \((r_1,r_2,r_3)\), each term in \(\mathcal{L}(W)\) is Tucker with rank at most \((r_1,r_2,r_3)\). Implementing:
\begin{itemize}[leftmargin=*]
    \item \texttt{applyL\_tucker(W, Dx, Dy, Dz, tol)} using Tucker arithmetic and \texttt{tucker\_sum},
    \item a Step-and-Truncate solver with Taylor order \(p\) and tolerance \((\Delta t)^{p+1}\),
    \item rank-vs-time tracking and error convergence for different \(\Delta t\).
\end{itemize}

\end{document}
